---
title: "Cotton&CottonSeed-Prices"
author: "Ananya, Hannah, Ekansh"
format: pdf
editor: visual
---

# Analysis of Cotton and Cottonseed Prices (1910-1938)

## Introduction

This report explores the historical trends in cotton and cottonseed prices from 1910 to 1938. The primary objectives are to identify significant trends, analyze price distributions, and uncover correlations between cotton and cottonseed prices.

## Data Description

The analysis uses the following datasets: 
- **Cotton Wholesale Prices (1910-1938)**: Contains annual wholesale prices of cotton. 
- **Cottonseed Prices (1910-1938)**: Provides annual prices for cottonseed. 
- **Merged Dataset**: Combines the above datasets for comparative analysis.
```{r}
## Loading the Data
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(readr)

# Load the datasets
cotton_prices <- read_csv("data\Cotton_Wholesale_Prices_1910_1938.csv")
cottonseed_prices <- read_csv("data\Cottonseed_Prices_1910_1938.csv")
merged_data <- read_csv("data\Merged_Cotton_Cottonseed_Prices.csv")
```

## Data Summary
```{r}
summary(cotton_prices)
summary(cottonseed_prices)
summary(merged_data)

```

## Exploratory Data Analysis
### Price Trends Over Time
```{r}
# Line plot for cotton prices
ggplot(merged_data, aes(x = Year, y = Cotton)) +
  geom_line(color = "blue", size = 1) +
  ggtitle("Cotton Prices (1910-1938)") +
  xlab("Year") +
  ylab("Price (USD)")

# Line plot for cottonseed prices
ggplot(merged_data, aes(x = Year, y = Cottonseed)) +
  geom_line(color = "green", size = 1) +
  ggtitle("Cottonseed Prices (1910-1938)") +
  xlab("Year") +
  ylab("Price (USD)")

```

### Distribution of Prices
```{r}
# Boxplot for both prices
merged_data_long <- merged_data %>%
  pivot_longer(cols = c(Cotton, Cottonseed), names_to = "Commodity", values_to = "Price")

ggplot(merged_data_long, aes(x = Commodity, y = Price, fill = Commodity)) +
  geom_boxplot() +
  ggtitle("Price Distribution: Cotton vs Cottonseed") +
  xlab("Commodity") +
  ylab("Price (USD)")

```

### Density Plots
```{r}
# Density plot for prices
ggplot(merged_data_long, aes(x = Price, fill = Commodity)) +
  geom_density(alpha = 0.5) +
  ggtitle("Density Plot: Cotton and Cottonseed Prices") +
  xlab("Price (USD)") +
  ylab("Density")

```

### Scatter Plots & Correlation
```{r}
# Scatter plot of cotton vs cottonseed prices
ggplot(merged_data, aes(x = Cotton, y = Cottonseed)) +
  geom_point(color = "purple", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  ggtitle("Relationship Between Cotton and Cottonseed Prices") +
  xlab("Cotton Price (USD)") +
  ylab("Cottonseed Price (USD)")

# Calculate correlation
cor(merged_data$Cotton, merged_data$Cottonseed, use = "complete.obs")
```

## INSIGHTS
1. Price Trends:

  - Cotton and cottonseed prices generally exhibit increasing trends with periodic fluctuations.
  - Seasonal changes might influence these patterns.
  
2. Price Variability:

  - Cotton prices are more volatile compared to cottonseed prices.
  
3. Relationship:

  - A positive correlation exists between cotton and cottonseed prices, indicating interdependence.
  
## Conclusion
This analysis demonstrates historical price trends and relationships between cotton and cottonseed markets. The findings can help in understanding market dynamics during the early 20th century.
